---
title: "STATSPROJECT"
author: "KIRAN RANGWANII"
date: "2023-04-14"
output:
  word_document: default
  html_document: default
---
##Reading data

##For analysis Perspective, better to go ahead with shorter Names only. 
# colnames(dataframe) <- c("Income", "Age","White recode",
#                          "Black or African American recode","Marital status","SEX","Educational attainment",
#                          "State","English Speaking Ability","Pacific Islander recode","Native Hawaiian recode",
#                          "American Indiana Alaska Native",
#                          "other race recode","Asian recode","Standard Occupational Classification")

```{r}
dataframe = read.csv("RAW-DATA.csv")
#dataframe = read.csv("tranformed_v1.csv")
colnames(dataframe)
```


```{r}
dim(dataframe)  
colnames(dataframe)
dataframe <- subset(dataframe, select = -X)
```
###Data cleaning
Making the data more efficient for our analysis

```{r}
dataframe<-subset(dataframe, POWSP!="N")
dim(dataframe)
dataframe <- subset(dataframe, POWSP < "072")
dim(dataframe)
dataframe<-subset(dataframe, SOCP!="N")
dim(dataframe)
dataframe <- subset(dataframe, PINCP >= 0)
```

###Recoding Marital Status to dummy variables
Married = 1 or else 0
Unmarried : rest of the categories as 1 and married as 0
```{r}
dataframe$married <- as.factor(ifelse(dataframe$MAR==1, 1, 0))

dataframe$unmarried <- as.factor(ifelse(dataframe$MAR %in% c(2,3,4,5), 1, 0))

dim(dataframe)
colnames(dataframe)

# class(dataframe$married)
# class(dataframe$unmarried)
```

```
###We do not recode ABILITY TO SPEAK ENGLISH to dummy variables as it's our variable of interest
$OnlyEnglish = 0
Very_well = 1
Well = 2
Not_well = 3
Not_at_all = 4
```
#Recoding EDUCATIONAL ATTAINMENT
##Considering values from 12th grade or higher, not considering values before 12th grade
##Here we choose Bachelors and above as one category
```{r}
# Create a new data frame with values from 16 to 24 in the SCHL column
dataframe <- subset(dataframe, SCHL >= 15 & SCHL <= 24 & SCHL!=17)
dataframe$NonGraduate <- as.factor(ifelse(dataframe$SCHL %in% c(15, 16,18,19,20),1,0))

dataframe$Graduate <-as.factor(ifelse(dataframe$SCHL %in% c(21,22,23,24),1,0))

dim(dataframe)
colnames(dataframe)

```
##Occupation recode to ddummy variables
##We group records that fall under same unique code
```{r}
dataframe$Managers=     ifelse(grepl("^11", dataframe$SOCP),1,0)
dataframe$Business=     ifelse(grepl("^131", dataframe$SOCP),1,0)
dataframe$Financial=   		as.factor(ifelse(grepl("^132", dataframe$SOCP),1,0))
dataframe$IT=          		as.factor(ifelse(grepl("^15", dataframe$SOCP),1,0))
dataframe$Engineering= 		as.factor(ifelse(grepl("^17", dataframe$SOCP), 1,0))
dataframe$Science=     		as.factor(ifelse(grepl("^19", dataframe$SOCP), 1,0))
dataframe$Coucelling=  		as.factor(ifelse(grepl("^21", dataframe$SOCP), 1,0))
dataframe$Legal=       		as.factor(ifelse(grepl("^23", dataframe$SOCP), 1,0))
dataframe$Education=   		as.factor(ifelse(grepl("^25", dataframe$SOCP), 1,0))
dataframe$Entertainment=	as.factor(ifelse(grepl("^27", dataframe$SOCP), 1,0))
dataframe$Medical=     		as.factor(ifelse(grepl("^29", dataframe$SOCP), 1,0))
dataframe$HealthCare=  		as.factor(ifelse(grepl("^31", dataframe$SOCP), 1,0))
dataframe$PRT=         		as.factor(ifelse(grepl("^33", dataframe$SOCP), 1,0))
dataframe$EATERY=      		as.factor(ifelse(grepl("^35", dataframe$SOCP), 1,0))
dataframe$CLN=         		as.factor(ifelse(grepl("^37", dataframe$SOCP), 1,0))
dataframe$PRS=         		as.factor(ifelse(grepl("^39", dataframe$SOCP), 1,0))
dataframe$Sales=       		as.factor(ifelse(grepl("^41", dataframe$SOCP), 1,0))
dataframe$Official=    		as.factor(ifelse(grepl("^43", dataframe$SOCP), 1,0))
dataframe$FFF=         		as.factor(ifelse(grepl("^45", dataframe$SOCP), 1,0))
dataframe$CON=         		as.factor(ifelse(grepl("^474", dataframe$SOCP),1,0))
dataframe$EXT=         		as.factor(ifelse(grepl("^475", dataframe$SOCP),1,0))
dataframe$RPR=         		as.factor(ifelse(grepl("^49", dataframe$SOCP), 1,0))
dataframe$Production=  		as.factor(ifelse(grepl("^51", dataframe$SOCP), 1,0))
dataframe$TRN=         		as.factor(ifelse(grepl("^53", dataframe$SOCP), 1,0))
dataframe$MIL=         		as.factor(ifelse(grepl("^55", dataframe$SOCP), 1,0))
dim(dataframe)
colnames(dataframe)
class(dataframe$MIL)
```
##As for state variable, we will only retain 50 states of the United States
##State
##072,166,251,254,301,303,399,555 are other countries hence we do not select those

```{r}
us_states <- sprintf("%03d", 001:056)

dataframe <- subset(dataframe, POWSP %in% us_states)
colnames(dataframe)
```
##This will form our population. As for our sample we will select East code states and create dummy variables for those.
#We will perform this part of recoding further down the project timeline



## Converting integer to discrete values 

```{r}
dataframe$ENG <- as.factor(dataframe$ENG)
dataframe$SEX <- as.factor(dataframe$SEX)
class(dataframe$ENG)
class(dataframe$SEX)
```

##Dependent variable log transformation
Log transformation performed because taking the logarithm of income can help to reduce the impact of these extreme values and make the distribution of income more symmetrical and easier to work with.
```{r}
colnames(dataframe)
hist(dataframe$PINCP,main="Before log Transformation",col = "RED")
dataframe$transformed_income<-log(dataframe$PINCP+1)
hist(dataframe$transformed_income,main="After log Transformation",col = "BLUE")
colnames(dataframe)

write.csv(dataframe,file = "tranformed_v1.csv",row.names = FALSE)
```


```{r}
#colnames(dataframe)
# hist(dataframe$PINCP,main="Before log Transformation",col = "RED")
# dataframe$transformed_income<-log(dataframe$PINCP+1)
# hist(dataframe$transformed_income,main="After log Transformation",col = "BLUE")
# colnames(dataframe)
# # class(dataframe$Production)
class(dataframe$ENG)
#write.csv(dataframe,file = "tranformed.csv",row.names = FALSE)
```

```{r}
#transformed_df = read.csv("tranformed.csv")
# class(transformed_df$ENG)
# class(transformed_df$SEX)
class(dataframe$Production)
class(dataframe$married)
#names(transformed_df)
```


```{r}
dataframe = read.csv("tranformed_v1.csv")
```
```{r}
colnames(dataframe)
```
##Checking for NA values
```{r}
sum(is.na(dataframe))
dataframe <- na.omit(dataframe)
nrow(dataframe)



```

substituting NA values with average value of column

```{r}
avg_value <- mean(dataframe$unmarried, na.rm = TRUE)
dataframe$unmarried[is.na(dataframe$unmarried)] <- avg_value
sum(is.na(dataframe$unmarried)) 

```
##summary of income transformation variable
```{r}
summary(dataframe$transformed_income)


```
```{r}
plot(ENG ~ transformed_income, data = dataframe)
plot(Engineering ~ transformed_income, data=dataframe)


```
```{r}
dataframe <- dataframe[complete.cases(dataframe[c("unmarried", "Bachelors")]), ]
write.csv(dataframe, "tranformed_v1.csv", row.names = FALSE)
dataframe = read.csv("tranformed_v1.csv")
dataframe <- na.omit(dataframe)

```
```{r}
library(ggplot2)
library(reshape2)
library(dplyr)
```


```{r}
#library(plotly)
dataframe$ENG <- as.factor(dataframe$ENG)
dataframe$SEX <- as.factor(dataframe$SEX)
dataframe$Bachelors <- as.factor(dataframe$Bachelors)
class(dataframe$ENG)
class(dataframe$SEX)
class(dataframe$Bachelors)
```
```{r}
dataframe$English <- as.factor(ifelse(transformed_df$ENG == 0, "Only English",
                                                  ifelse(transformed_df$ENG == 1, "Very well",
                                                         ifelse(transformed_df$ENG == 2, "Well",
                                                         ifelse(transformed_df$ENG == 3, "Not Well",
                                                                ifelse(transformed_df$ENG == 4, "Not Not at all",
                                                                       "Not at all"))))))
#names(transformed_df)

dataframe$Gender <- as.factor(ifelse(transformed_df$SEX == 1, "Male",
                                                  ifelse(transformed_df$SEX == 2, "Female","no")))
```
```{r}
income_eng <- dataframe[, c("PINCP", "English")]

```
```{r}
income_eng_mean_median <- income_eng %>%
  group_by(English) %>%
  summarize(mean_income = mean(PINCP), median_income = median(PINCP))
income_eng_mean_median
```
```{r}
ggplot(income_eng_mean_median, aes(x = English, y = median_income, fill = English)) +
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "English Proficiency") +
  scale_x_discrete(labels = c("Only English", "Very well", "Well",
                              "Not well", "Not at all")) +
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  )+
   labs(title = "Median Income by English Proficiency")+
   xlab("Ability to Speak English")+
   ylab("Median Income ($)")

ggplot(income_eng_mean_median, aes(x = English, y = mean_income, fill = English)) +
  geom_bar(stat = "identity") +
  scale_fill_discrete("English Proficiency") +
  scale_x_discrete(labels = c("Only English", "Very well", "Well",
                              "Not well", "Not at all")) +
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  )+
  labs(title = "Mean Income by English Proficiency")+
  xlab("Ability to Speak English")+
  ylab("Median Income ($)")
```
```{r}
income_grad <- dataframe[, c("PINCP", "Bachelors")] 

income_graduation <- income_grad %>%
  group_by(Bachelors) %>%
  summarize(mean_income = mean(PINCP), median_income = median(PINCP))
income_graduation
```
```{r}
class(income_graduation$Bachelors)
```
```{r}
income_graduation <- income_grad %>%
  group_by(Bachelors) %>%
  summarize(mean_income = mean(PINCP), median_income = median(PINCP))

income_graduation$Bachelors <- factor(income_graduation$Bachelors, levels = c("0", "1"), labels = c("Less than Bachelor", "Bachelor's or more"))

ggplot(income_graduation, aes(x = Bachelors, y = median_income, fill = Bachelors)) +
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "Education Level")+
  scale_x_discrete(labels = c("No Bachelor's", "Bachelor's or more"))+
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  ) +
  labs(title = "Median Income by Educational Attainment",
       fill = "Graduate")+
   xlab("Degree Holding Status")+
   ylab("Median Income ($)")


ggplot(income_graduation, aes(x = Bachelors, y = mean_income, fill = Bachelors)) +
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "Education Level") +
  scale_x_discrete(labels = c("No Bachelor's", "Bachelor's or more"))+
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  )+
  labs(title = "Mean Income by Educational Attainment",
       fill = "Bachelors")+
   xlab("Degree Holding Status")+
   ylab("Median Income ($)")
```

```{r}
income_marital_status <- dataframe[, c("PINCP", "married")] 

income_marital_status <- income_marital_status %>%
  group_by(married) %>%
  summarize(mean_income = mean(PINCP), median_income = median(PINCP))
income_marital_status$married <- factor(income_marital_status$married, levels = c("0", "1"), labels = c("Single", "Married"))
income_marital_status
```

```{r}
ggplot(income_marital_status, aes(x = married, y = mean_income, fill = married)) +
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "Marital Status") +
  scale_x_discrete(labels = c("Single", "Married")) +
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  )+
  labs(title = "Mean Income by Marital Status")+
  xlab("Marital Status")+
  ylab("Median Income ($)")


ggplot(income_marital_status, aes(x = married, y = median_income, fill = married)) +
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "Marital Status") +
  scale_x_discrete(labels = c("Single", "Married")) +
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  )+
  labs(title = "Median Income by Marital Status")+
  xlab("Marital Status")+
  ylab("Median Income ($)")
```
```{r}
income_gender <- dataframe[, c("PINCP", "Gender")] 

income_graduation <- income_gender %>%
  group_by(Gender) %>%
  summarize(mean_income = mean(PINCP), median_income = median(PINCP))
income_graduation
```
```{r}
ggplot(income_graduation, aes(x = Gender, y = median_income, fill = Gender)) +
  geom_bar(stat = "identity") +
  scale_fill_discrete() +
  scale_x_discrete(labels = c("Female", "Male")) +
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  )+
  labs(title = "Median Income by Gender")+
  xlab("Gender")+
  ylab("Median Income ($)")
```
```{r}
ggplot(dataframe, aes(x = dataframe$AGE, y = dataframe$PINCP, color = dataframe$AGE)) + 
  geom_point() + 
  labs(title = "Age vs. Income", x = "Age", y = "Income")
```
```{r}
subset_data_profession <- dataframe[, c("PINCP", "Managers", "Business", "Financial", "IT", "Engineering", "Science", "Coucelling", "Legal", "Education", "Entertainment", "Medical", "HealthCare", "PRT", "EATERY", "CLN", "PRS", "Sales", "Official", "FFF", "CON", "EXT", "RPR", "Production", "TRN", "MIL")]

subset_melted_data <- melt(subset_data_profession,id.vars = "PINCP", variable.name = "profession")
subset_melted_data <-subset(subset_melted_data,value == 1)
subset_melted_data <- subset_melted_data %>% rename(income = PINCP)
subset_melted_data
income_by_profession <- aggregate(income ~ profession, subset_melted_data, median)
```
```{r}
ggplot(income_by_profession, aes(x = profession, y = income, fill = profession)) +
  geom_bar(stat = "identity") +
  scale_fill_discrete(name = "Profession") +
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text.y = element_text(angle = 0, hjust = 1, size = 8), # change the size of y-axis text
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  ) +
  coord_flip() +
  labs(title = "Median Income by Professions", x = "Professions", y = "Median Income($)")
```
```{r}
subset_data_race <- dataframe[, c("PINCP", "RACWHT", "RACBLK", "RACPI", "RACNH","RACAIAN","RACSOR","RACASN")]
subset_melted_data_race <- melt(subset_data_race,id.vars = "PINCP", variable.name = "race")
subset_melted_data_race <-subset(subset_melted_data_race,value == 1)
#income_by_race <- aggregate(PINCP ~ race, subset_melted_data_race, mean)
subset_melted_data_race <- subset_melted_data_race %>% rename(income = PINCP)
subset_melted_data_race
income_by_race <- aggregate(income ~ race, subset_melted_data_race, median)
```
```{r}
ggplot(income_by_race, aes(x = race, y = income, fill = race)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = c("White", "Black or African American","Pacific Islanders","Native Hawaiian","Native Alaskan","Other","Asian" )) +
  scale_fill_discrete(name = "Race",labels = c("White", "Black or African American","Pacific Islanders","Native Hawaiian","Native Alaskan","Other","Asian" ))  +
  theme(
    panel.background = element_rect(fill = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "right"
  )+
  labs(title = "Median Income by Race")+
  xlab("Race")+
  ylab("Median Income ($)")
```


#MODEL 1 WITH ALL VARIABLES

Removing married and NoBachelors as we don't need those{Unmarried and Bachelors cover all the values}

```{r}
model <- lm(transformed_income ~ ENG+AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP +Bachelors+unmarried+ Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL, data = dataframe,na.action=na.exclude)


```


```{r}
summary(model)
confint(model)
anova(model)
```

The R squared value is 0.3426,indicates that 34.26% of the variation in income  is explained by or associated with independent variables. The adjusted R-squared of 0.3426 indicates that 34.26% of the variation in income is explained by independent  after adjusting for the number of independent variables. 
```{r}
par(mfrow=c(2,2))
plot(model)
```

```{r}
plot(model, which = 1)

```
```{r}
install.packages("coefplot")
library(coefplot)
```



```{r}
coefplot(model,color= )
```





```{r}
#install.packages("margins")

#library(margins)
#marginal_effects <- margins(model, variables = "ENG")
#plot(marginal_effects)

```
#MODEL WITHOUT VARIABLE OF INTEREST : ENGLISH

```{r}
model_without_eng <- lm(transformed_income ~ AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP + unmarried + Bachelors + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL, data = dataframe,na.action=na.exclude)
```
```{r}
summary(model_without_eng)
confint(model_without_eng)
anova(model_without_eng)
```
The R squared and adjusted R-squared increased a bit, thus it can be said that ENG has negligible but some impact on income

##MODEL WITHOUT RACE
```{r}
model_without_RACE <- lm(transformed_income ~ ENG+AGEP + SEX + POWSP + unmarried+ Bachelors + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL, data = dataframe,na.action=na.exclude)
```


```{r}
summary(model_without_RACE)

confint(model_without_RACE)
anova(model_without_RACE)

```
##MODEL WITHOUT OCCUPATION
```{r}
model_without_occup <- lm(transformed_income ~ ENG+AGEP + SEX + POWSP  +RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH +unmarried + Bachelors, data = dataframe,na.action=na.exclude)

```
```{r}
summary(model_without_occup)
confint(model_without_occup)
anova(model_without_occup)

```
##MODEL WITHOUT MARRAIGE
```{r}
model_without_mar <- lm(transformed_income ~ AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP  + Bachelors + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL+ENG, data = dataframe,na.action=na.exclude)

```
```{r}
summary(model_without_mar)
confint(model_without_mar)
anova(model_without_mar)


```
##MODEL WITHOUT EDUCATION
```{r}
model_without_edu <- lm(transformed_income ~ AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP   + unmarried  + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL+ENG, data = dataframe,na.action=na.exclude)

```
```{r}
summary(model_without_edu)
confint(model_without_edu)
anova(model_without_edu)


```
##MODEL WITH ONLY ENGLISH
```{r}
model_only_eng <- lm(transformed_income ~ENG , data = dataframe,na.action=na.exclude)
summary(model_only_eng)
confint(model_only_eng)
anova(model_only_eng)

```
##MODEL WITH ONLY FEW OCCUPATIONS
```{r}
model_only_few_occu <- lm(transformed_income ~ ENG+AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP  + unmarried  + Bachelors + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + Sales , data = dataframe,na.action=na.exclude)

summary(model_only_few_occu)
confint(model_only_few_occu)
anova(model_only_few_occu)

```
##EAST COAST STATE SUBSET 
```{r}
state_subset_df<- dataframe[dataframe$POWSP %in% c("23", "33", "25", "44", "9","36","34","10","24","51","37","45", "13","12"), ]

state_subset_df <- na.omit(state_subset_df)

colnames(state_subset_df)
nrow(state_subset_df)


```
```{r}
model_state_sample <- lm(transformed_income ~ ENG+AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP+unmarried+Bachelors+ Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL, data = state_subset_df)

```
```{r}
summary(model_state_sample)
confint(model_state_sample)
anova(model_state_sample)
```

