---
title: "STATSPROJECT"
author: "KIRAN RANGWANII"
date: "2023-04-14"
output:
  word_document: default
  html_document: default
---
##Reading data

##For analysis Perspective, better to go ahead with shorter Names only. 
# colnames(dataframe) <- c("Income", "Age","White recode",
#                          "Black or African American recode","Marital status","SEX","Educational attainment",
#                          "State","English Speaking Ability","Pacific Islander recode","Native Hawaiian recode",
#                          "American Indiana Alaska Native",
#                          "other race recode","Asian recode","Standard Occupational Classification")

```{r}
dataframe = read.csv("RAW-DATA.csv")
#dataframe = read.csv("tranformed_v1.csv")
colnames(dataframe)
```


```{r}
dim(dataframe)  
colnames(dataframe)
dataframe <- subset(dataframe, select = -X)
```
###Data cleaning


```{r}
dataframe<-subset(dataframe, POWSP!="N")
dim(dataframe)
dataframe <- subset(dataframe, POWSP < "072")
dim(dataframe)
dataframe<-subset(dataframe, SOCP!="N")
dim(dataframe)
dataframe <- subset(dataframe, PINCP >= 0)
```

###Recoding Marital Status to dummy variables

```{r}
dataframe$married <- as.factor(ifelse(dataframe$MAR==1, 1, 0))

dataframe$unmarried <- as.factor(ifelse(dataframe$MAR %in% c(2,3,4,5), 1, 0))

dim(dataframe)
colnames(dataframe)

# class(dataframe$married)
# class(dataframe$unmarried)
```

```
###We do not recode ABILITY TO SPEAK ENGLISH to dummy variables as it's our variable of interest
$OnlyEnglish = 0
Very_well = 1
Well = 2
Not_well = 3
Not_at_all = 4
```
#Recoding EDUCATIONAL ATTAINMENT
##Considering values from 12th grade or higher, not considering values before 12th grade
##Here we choose Bachelors and above as one category
```{r}
# Create a new data frame with values from 16 to 24 in the SCHL column
dataframe <- subset(dataframe, SCHL >= 15 & SCHL <= 24 & SCHL!=17)
dataframe$NonGraduate <- as.factor(ifelse(dataframe$SCHL %in% c(15, 16,18,19,20),1,0))

dataframe$Graduate <-as.factor(ifelse(dataframe$SCHL %in% c(21,22,23,24),1,0))

dim(dataframe)
colnames(dataframe)

```
##Occupation recode to ddummy variables
##We group records that fall under same unique code
```{r}
dataframe$Managers=     ifelse(grepl("^11", dataframe$SOCP),1,0)
dataframe$Business=     ifelse(grepl("^131", dataframe$SOCP),1,0)
dataframe$Financial=   		as.factor(ifelse(grepl("^132", dataframe$SOCP),1,0))
dataframe$IT=          		as.factor(ifelse(grepl("^15", dataframe$SOCP),1,0))
dataframe$Engineering= 		as.factor(ifelse(grepl("^17", dataframe$SOCP), 1,0))
dataframe$Science=     		as.factor(ifelse(grepl("^19", dataframe$SOCP), 1,0))
dataframe$Coucelling=  		as.factor(ifelse(grepl("^21", dataframe$SOCP), 1,0))
dataframe$Legal=       		as.factor(ifelse(grepl("^23", dataframe$SOCP), 1,0))
dataframe$Education=   		as.factor(ifelse(grepl("^25", dataframe$SOCP), 1,0))
dataframe$Entertainment=	as.factor(ifelse(grepl("^27", dataframe$SOCP), 1,0))
dataframe$Medical=     		as.factor(ifelse(grepl("^29", dataframe$SOCP), 1,0))
dataframe$HealthCare=  		as.factor(ifelse(grepl("^31", dataframe$SOCP), 1,0))
dataframe$PRT=         		as.factor(ifelse(grepl("^33", dataframe$SOCP), 1,0))
dataframe$EATERY=      		as.factor(ifelse(grepl("^35", dataframe$SOCP), 1,0))
dataframe$CLN=         		as.factor(ifelse(grepl("^37", dataframe$SOCP), 1,0))
dataframe$PRS=         		as.factor(ifelse(grepl("^39", dataframe$SOCP), 1,0))
dataframe$Sales=       		as.factor(ifelse(grepl("^41", dataframe$SOCP), 1,0))
dataframe$Official=    		as.factor(ifelse(grepl("^43", dataframe$SOCP), 1,0))
dataframe$FFF=         		as.factor(ifelse(grepl("^45", dataframe$SOCP), 1,0))
dataframe$CON=         		as.factor(ifelse(grepl("^474", dataframe$SOCP),1,0))
dataframe$EXT=         		as.factor(ifelse(grepl("^475", dataframe$SOCP),1,0))
dataframe$RPR=         		as.factor(ifelse(grepl("^49", dataframe$SOCP), 1,0))
dataframe$Production=  		as.factor(ifelse(grepl("^51", dataframe$SOCP), 1,0))
dataframe$TRN=         		as.factor(ifelse(grepl("^53", dataframe$SOCP), 1,0))
dataframe$MIL=         		as.factor(ifelse(grepl("^55", dataframe$SOCP), 1,0))
dim(dataframe)
colnames(dataframe)
class(dataframe$MIL)
```
##As for state variable, we will only retain 50 states of the United States
##State
##072,166,251,254,301,303,399,555 are other countries hence we do not select those

```{r}
us_states <- sprintf("%03d", 001:056)

dataframe <- subset(dataframe, POWSP %in% us_states)
colnames(dataframe)
```
##This will form our population. As for our sample we will select East code states and create dummy variables for those.
#We will perform this part of recoding further down the project timeline



## Converting integer to discrete values 

```{r}
dataframe$ENG <- as.factor(dataframe$ENG)
dataframe$SEX <- as.factor(dataframe$SEX)
class(dataframe$ENG)
class(dataframe$SEX)
```

##Dependent variable log transformation

```{r}
#colnames(dataframe)
# hist(dataframe$PINCP,main="Before log Transformation",col = "RED")
# dataframe$transformed_income<-log(dataframe$PINCP+1)
# hist(dataframe$transformed_income,main="After log Transformation",col = "BLUE")
# colnames(dataframe)
# # class(dataframe$Production)
class(dataframe$ENG)
#write.csv(dataframe,file = "tranformed.csv",row.names = FALSE)
```

```{r}
#transformed_df = read.csv("tranformed.csv")
# class(transformed_df$ENG)
# class(transformed_df$SEX)
class(dataframe$Production)
class(dataframe$married)
#names(transformed_df)
```

```{r}
transformed_df$ENG
```

```{r}
dataframe = read.csv("tranformed_v1.csv")
```
```{r}
colnames(dataframe)
```
```{r}
sum(is.na(dataframe))
dataframe <- na.omit(dataframe)
nrow(dataframe)



```

```{r}
summary(dataframe$transformed_income)

```
```{r}
plot(ENG ~ transformed_income, data = dataframe)
plot(Engineering ~ transformed_income, data=dataframe)


```
```{r}
dataframe <- dataframe[complete.cases(dataframe[c("unmarried", "Bachelors")]), ]
write.csv(dataframe, "tranformed_v1.csv", row.names = FALSE)
dataframe = read.csv("tranformed_v1.csv")

```



```{r}
model <- lm(transformed_income ~ AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP +  married + unmarried + NoBachelors + Bachelors + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL+ENG, data = dataframe,na.action=na.exclude)

```


```{r}
summary(model)
```

The R squared value is 0.3426,indicates that 34.26% of the variation in income  is explained by or associated with independent variables. The adjusted R-squared of 0.3426 indicates that 34.26% of the variation in income is explained by independent  after adjusting for the number of independent variables. 
```{r}
par(mfrow=c(2,2))
plot(model)
```

```{r}
plot(model, which = 1)

```
```{r}
install.packages("coefplot")
library(coefplot)
coefplot(model)
```


```{r}
install.packages("margins")

library(margins)
marginal_effects <- margins(model, variables = "ENG")
plot(marginal_effects)

```


```{r}
model_without_eng <- lm(transformed_income ~ AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP  + married + unmarried + NoBachelors + Bachelors + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL, data = dataframe,na.action=na.exclude)
```
```{r}
summary(model_without_eng)
```
The R squared and adjusted R-squared increased a bit, thus it can be said that ENG has negligible but some impact on income
```{r}
model_without_RACE <- lm(transformed_income ~ ENG+AGEP + SEX + POWSP  + married + unmarried + NoBachelors + Bachelors + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL, data = dataframe,na.action=na.exclude)
```


```{r}
summary(model_without_RACE)
```

```{r}
model_without_occup <- lm(transformed_income ~ ENG+AGEP + SEX + POWSP + PWGTP +RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH+ married + unmarried + NoBachelors + Bachelors, data = dataframe,na.action=na.exclude)

```
```{r}
summary(model_without_occup)

```

```{r}
model_without_mar <- lm(transformed_income ~ AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP + PWGTP  + NoBachelors + Bachelors + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL+ENG, data = dataframe,na.action=na.exclude)

```
```{r}
summary(model_without_mar)

```
```{r}
model_without_edu <- lm(transformed_income ~ AGEP + SEX + RACWHT + RACAIAN + RACSOR + RACASN + RACBLK + RACNH + POWSP + PWGTP + married + unmarried  + Managers + Business + Financial + IT + Engineering + Science + Coucelling + Legal + Education + Entertainment + Medical + HealthCare + PRT + EATERY + CLN + PRS + Sales + Official + FFF + CON + EXT + RPR + Production + TRN + MIL+ENG, data = dataframe,na.action=na.exclude)

```
```{r}
summary(model_without_edu)

```
```{r}
model_only_eng <- lm(transformed_income ~ENG , data = dataframe,na.action=na.exclude)
summary(model_only_eng)
```

